---
title: "Week 1: Data Types and Structures"
author: "J.Blondeau"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Types and Structures in R {.tabset}

In this first lesson, we will explore the fundamental building blocks of data in R. A firm grasp of these concepts is the most important skill for troubleshooting code. Nearly every error you will encounter is a result of data not being in the expected format or structure. This lesson provides the foundation for effective data manipulation and analysis.

### Data Types

Everything you create in R has a "type." Think of it like classifying organisms; you can't work with them properly if you don't know what they are. The most common types you'll encounter are:

#### **Numeric**
Numeric data represents numbers, including integers (whole numbers) and floating-point numbers (decimals).

```{r}
# Numeric data examples
depth_m <- 15.5
fish_count <- 42
```

#### **Character**
Character data, also known as "strings," represents text. You must use quotation marks (`"` or `'`) to create them.

```{r}
# Character data examples
species_name <- "Acropora palmata"
site_id <- "STT-04"
```

#### **Logical (Boolean)**
Logical data represents binary `TRUE` or `FALSE` values. These are essential for filtering data. **Note:** They must be in all caps!

```{r}
# Logical data examples
is_bleached <- TRUE
transect_completed <- FALSE
```

#### **Why This Matters: A Practical Example**
What happens if you accidentally record data as the wrong type? Let's say we included units in our depth measurements.

```{r}
# A vector with a mix of numbers and one character string
depths <- c(10.2, 11.5, "9.8m", 10.5)

# Let's try to calculate the average depth...
# mean(depths) 
# This will throw an error! R can't do math on text.
# Uncomment the line above to see the error for yourself.
```
This is one of the most common errors in R! The fix is to make sure your data is the correct **type** before you try to analyze it.

---

### Inspecting Your Data

These functions will give you insight to the data types and structures. Use them constantly!

#### **`class()` Function**
The `class()` function tells you the specific data type or structure of an R object.

```{r}
# Using class() to determine the class of an object
class(depth_m)
class(species_name)
class(is_bleached)
```

#### **`str()` Function**
The `str()` (structure) function provides a concise summary of any R object's structure. It's incredibly useful for lists and data frames.

```{r}
# Let's create a list to hold info about a survey
# (We'll learn more about lists below!)
t1_analysis <- list(
  site_name = "Brewer's Bay West",
  survey_date = "2025-08-24",
  depths_m = c(8.1, 8.3, 8.2)
)

# str() gives us a perfect summary of what's inside
str(t1_analysis)
```

#### **`glimpse()` Function**
The `glimpse()` function from the `dplyr` package (part of the `tidyverse`) is the best way to get a quick, clean overview of a data frame.

```{r}
# Using glimpse() to get an overview of a data frame
library(dplyr)

# A more realistic data frame
fish_survey <- data.frame(
  transect_id = c("T1", "T1", "T2"),
  species = c("Stegastes partitus", "Acanthurus bahianus", "Stegastes partitus"),
  count = c(12, 5, 18)
)

glimpse(fish_survey)
```

---

### Data Structures

If data types are the atoms, data structures are the molecules. They are containers for holding your data.

#### **Vectors**
Vectors are the most fundamental structure. They are a sequence of elements of the **same data type**.

```{r}
# Vector examples
depths <- c(5.2, 5.5, 6.1, 5.8)
sites <- c("Coki Point", "Brewer's Bay", "Magens Bay")

## 1) What are the data types in each of the above vectors?
class(depths)
class(sites)
```

**The Vector's One-Rule Policy:** A vector can only hold one data type! If you try to mix them, R will quietly **coerce** them to match the most flexible type (usually character). This silent conversion can cause major headaches.

```{r}
mixed <- c(10, "Siderastrea", TRUE)

## 2) What do you think the data type of `mixed` will be?
class(mixed)
# It's "character"! R downgraded the number 10 to "10" and TRUE to "TRUE".
```

#### **Data Frames**
A data frame is the most important data structure for data analysis. It's a table, like you'd see in Excel.
* Each **column** is a vector, where all elements must be the same data type.
* Different columns can have different data types.

```{r}
# We already created this data frame earlier
glimpse(fish_survey)
```

#### **Lists**
Lists are the most flexible data structure. You can think of them as a **digital lab binder** or a project folder. A list can contain *anything*—vectors, data frames, even other lists! They are great for keeping related but dissimilar items together.

```{r}
# We already created this list earlier, which holds info for our T1 analysis
str(t1_analysis)
```

---

### Reading & Writing Data

Usually, our data lives in external files like `.csv` or `.xlsx`. A huge part of your work in R will be reading data in and writing results out. We'll use functions from the `tidyverse` for this.

**A Quick Note on File Paths:** The #1 reason reading data fails is an incorrect file path. The best practice is to use **RStudio Projects** (`.Rproj` files), which automatically sets the working directory to your project folder. Then you can use simple paths like `"data/my_file.csv"`.

#### **`read_csv()` Function**
The `read_csv()` function from the `readr` package is the best way to read data from CSV (Comma Separated Values) files. It's fast, reliable, and won't do weird things to your data like the old base R functions sometimes do.

```{r message=FALSE, warning=FALSE}
# Reading CSV file using read_csv()
library(readr)

# Assuming you have a folder named "data" in your R Project
# boat_logs <- read_csv("data/sttstj_2023_boat_logs.csv")

# head(boat_logs)
```

#### **`read_xlsx()` Function**
The `read_xlsx()` function from the `readxl` package allows you to read data from Excel files. You can (and should!) specify the sheet name.

```{r}
library(readxl)

#xlsx_data <- read_xlsx("data/sttstj_all_sites.xlsx", sheet = "locations")

## 1) What would happen if you removed the 'sheet =' argument?
## Answer: It would read the first sheet by default. It's always safer to be specific!

#head(xlsx_data)
```

#### **`write_csv()` & `write_xlsx()`**
These functions do the opposite: they save your R objects (like data frames) to files that can be shared or opened in other programs.

```{r}
# Exporting our fish_survey data frame to a CSV
# write_csv(fish_survey, "data/fish_survey_output.csv")

# The writexl package can write a list of data frames to different sheets
library(writexl)
# write_xlsx(list("Fish Survey" = fish_survey), "data/survey_output.xlsx")
```

---

### Case Study: The Silent Error

Let's tie everything together. You've been given data on coral bleaching severity, measured on a 1-5 scale. You want to find the average severity.

```{r}
# Your data vector
bleaching_scores <- c(3, 4, 2, 5, "Not Assessed", 3)

# 1. The Goal: Calculate the mean bleaching score
# mean(bleaching_scores) 
# This gives an error! Why?

# 2. The Investigation:
class(bleaching_scores)
# It's a "character" vector. The presence of the text "Not Assessed"
# caused R to coerce the entire vector to character type.

# 3. The Solution: We need to clean the data *before* we analyze it.

```
